---
author: "Gabriele Tomè"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmdformats::readthedown:
    df_print: paged
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
---

# Tester PhosProTN package

```{r load functions, warning=F, message=FALSE, results='hide'}
# setwd("/home/gabriele/Desktop/UNITN/proTN_package/")
outputdir = "results_phospho_proteome"
suppressPackageStartupMessages(suppressMessages(suppressWarnings({
  lapply(list.files("./R", pattern = ".R", full.names = T), function(x){if(!file.info(x)$isdir){source(x)}})
  
  # Package requided
  library(data.table)
  library(readxl)
  library(stringi)
  library(ggplot2)
  library(DEqMS)
  library(proBatch)
  library(writexl)
  library(ggrepel)
  library(ggbeeswarm)
  library(limma)
  library(plotly)
  library(enrichR)
  library(qpdf)
  library(STRINGdb)
  library(PhosR)
  
  library(pheatmap)
  library(svgPanZoom)
  library(SummarizedExperiment)
  library(parallel)
})))
```


# Test proteomics pipeline

```{r read_proteomics_MQ}
proteome_data <- read_phospho_proteome_proteomics(software = "MQ", 
                                                  folder_proteome = "./data/case_study_phospho/txt_PROTEOME/",
                                                  folder_phospho = "./data/case_study_phospho/txt_Phospho/",
                                                  peptide_proteome_filename = "evidence", 
                                                  peptide_phospho_filename = "evidence", 
                                                  phospho_thr = 0.75)
```

```{r read_proteomics_PD, eval=FALSE}
proteome_data <- read_phospho_proteome_proteomics(software = "PD", folder = "data/PD/phospho", annotation_filename = "INPUT", proteinGroup_filename = "PROT.xlsx")
```

```{r preprocess_figure}
generate_abundance <- generate_abundance_plot(proteome_data = proteome_data, phospho_with_proteome=TRUE)
generate_abundance

generate_peptide_distribution <- generate_peptide_distribution_plot(proteome_data = proteome_data,
                                                                    phospho_with_proteome=TRUE)
generate_peptide_distribution

create_phosphosite_fig <- create_phosphosite_plot(proteome_data = proteome_data, 
                                                  software = "MQ")
create_phosphosite_fig
```

# Test impute function

```{r impute_function}
imputed_data <- impute_intensity(proteome_data = proteome_data)
imputed_data
```

# Test normalization

```{r normalization_function}
normalized_data <- normalization_ProTN(proteome_data = imputed_data)
normalized_data
```

# Test Batch Correction -- NOT TESTED YET

```{r batch_function, eval=FALSE}
# TODO: NEED TO TEST
batch_coor_data <- batch_correction(proteome_data = normalized_data, batch_col = "batch")
batch_coor_data
```

# Test Exporter

```{r exporter, eval=FALSE}
save_abundance_tables(proteome_data = normalized_data, dirOutput = outputdir)
```

# Test figure distribution normalized

```{r normalized_plot}
prot_plot <- plot_abundance_distribution(proteome_data = normalized_data, type = "protein")
prot_plot
pep_plot <- plot_abundance_distribution(proteome_data = normalized_data, type = "peptide")
pep_plot
```

# MDS AND PCA

```{r mds_normalized}
prot_mds <- mds_plot(proteome_data = normalized_data, type = "protein")
prot_mds
pep_mds <- mds_plot(proteome_data = normalized_data, type = "peptide")
pep_mds
```

```{r pca_normalized}
prot_mds <- pca_plot(proteome_data = normalized_data, type = "protein")
prot_mds
pep_mds <- pca_plot(proteome_data = normalized_data, type = "peptide")
pep_mds
```

# List protein boxplot

```{r boxplot_protein}
prot_boxplot <- plot_selected_proteins(proteome_data = normalized_data, list_protein = c("Aaas","Aacs","Zyx"))
prot_boxplot
```

# Differential analysis

```{r differential_analysis}
formule_contrast <- list("A2016vsWT"="A2016T-wt")

differential_results <- differential_analysis(proteome_data = normalized_data, formule_contrast = formule_contrast, 
                                              fc_thr=0.75, pval_fdr = "p_val", pval_thr=0.05, signal_thr=0)

differential_results

save_differential_analysis_table(proteome_data = normalized_data, differential_results = differential_results, 
                                 dirOutput=outputdir)

ploft_diff_number_ctrl <- generate_differential_barplots(differential_results, data_type="protein", 
                                                         phospho_ctrl = T, color_contrast = c("grey40","grey60",
                                                                                              "grey40","grey60",
                                                                                              "#664069","#8A628D"))
ploft_diff_number_ctrl

ploft_diff_number <- generate_differential_barplots(differential_results, data_type="peptide")
ploft_diff_number
```

# Test vulcano plot

```{r vulcano_plot}
generate_volcano_plots_peptide<-generate_volcano_plots(differential_results, data_type="peptide",
                                                       comparison="A2016vsWT")
generate_volcano_plots_peptide
```


# MDS differential analysis

```{r mds_differential}
mds_diff_pep <- mds_differential_analysis_plot(differential_analysis = differential_results, 
                                               proteome_data = normalized_data, 
                                               type = "peptide")
mds_diff_pep
```

# pca differential analysis

```{r pca_differential}
pca_diff_pep <- pca_differential_analysis_plot(differential_analysis = differential_results, proteome_data = normalized_data, type = "peptide")
pca_diff_pep
```

# TEST enrichement

```{r enrichment}
enrichemnt_results <- perform_enrichment_analysis(differential_results = differential_results, 
                                                  enrichR_custom_DB = T,
                                                  enrich_filter_DBs=c("Human_Gene_Atlas",
                                                                      "BioPlanet_2019",
                                                                      "KEGG_2021_Human",
                                                                      "Reactome_2022",
                                                                      "ChEA_2022"), 
                                                  dirOutput = outputdir)

plots_all <- enrichment_figure(enr_df = enrichemnt_results, category = c("all"), save=T, dirOutput = outputdir)
plots_all

plots_down <- enrichment_figure(enr_df = enrichemnt_results, category = c("down","up"), 
                                save=F, color_contrast = c("grey40","grey60"))
plots_down

plots_terms <- enrichment_figure(enr_df = enrichemnt_results, category = c("down","up"), save=F, enrich_filter_term = c("cell","pathways"))
plots_terms
```

# TEST UNIVERSE ENRICHMENT

```{r universe_enrichment}
# TODO: add universe enrichment
```

# TEST network string

```{r string network}

strindb_res <- STRINGdb_network(differential_results = differential_results, species="Homo sapiens", dirOutput = outputdir)
strindb_res  

```

# TEST KINASE TREE

```{r kinase_tree}
kinase_res <- kinase_tree(proteome_data = normalized_data, 
                          differential_results = differential_results, 
                          formule_CORAL = formule_contrast, 
                          dirOutput = outputdir)
kinase_res
```

