---
author: "Gabriele Tomè"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmdformats::readthedown:
    df_print: paged
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
---

# Tester PhosProTN package

```{r set_dir, warning=F, message=FALSE, results='hide'}
# setwd("/home/gabriele/Desktop/UNITN/proTN_package/")
outputdir = "results_phospho"
devtools::load_all(".")
```

# Test proteomics pipeline

```{r read_proteomics_MQ, eval=FALSE}

proteome_data <- read_phosphoproteomics(software = "MQ",
                                        folder = "inst/extdata/case_study_phospho/txt_Phospho", 
                                        peptide_filename = "evidence",
                                        phospho_thr = 0.75)
```

```{r read_proteomics_PD}
proteome_data <- read_phosphoproteomics(software = "PD",
                                        folder = "inst/extdata/PD/phospho",
                                        annotation_filename = "INPUT",
                                        proteinGroup_filename = "PROT.xlsx")
```

```{r preprocess_figure}
generate_abundance <- generate_abundance_plot(proteome_data = proteome_data)
generate_abundance

generate_peptide_distribution <- generate_peptide_distribution_plot(proteome_data = proteome_data)
generate_peptide_distribution

create_phosphosite_fig <- create_phosphosite_plot(proteome_data = proteome_data,
                                                  software = "PD")
create_phosphosite_fig
```

# Test impute function

```{r impute_function}
imputed_data <- impute_intensity(proteome_data = proteome_data)
```

# Test normalization

```{r normalization_function}
normalized_data <- normalization_ProTN(proteome_data = imputed_data)
```

# Test Batch Correction -- NOT TESTED YET

```{r batch_function, eval=FALSE}
# TODO: NEED TO TEST
batch_coor_data <- batch_correction(proteome_data = normalized_data,
                                    batch_col = "batch")
```

# Test Exporter

```{r exporter}
save_abundance_tables(proteome_data = normalized_data, 
                      dirOutput=outputdir)
```

# Test figure distribution normalized

```{r normalized_plot}
prot_plot <- plot_abundance_distribution(proteome_data = normalized_data,
                                         type = "protein")
prot_plot
pep_plot <- plot_abundance_distribution(proteome_data = normalized_data,
                                        type = "peptide")
pep_plot
```

# MDS AND PCA

```{r mds_normalized}
prot_mds <- mds_plot(proteome_data = normalized_data,
                     type = "protein")
prot_mds
pep_mds <- mds_plot(proteome_data = normalized_data,
                    type = "peptide")
pep_mds
```

```{r pca_normalized}
prot_mds <- pca_plot(proteome_data = normalized_data,
                     type = "protein")
prot_mds
pep_mds <- pca_plot(proteome_data = normalized_data,
                    type = "peptide")
pep_mds
```

# List protein boxplot

```{r boxplot_protein}
prot_boxplot <- plot_selected_proteins(proteome_data = normalized_data, 
                                       list_protein = c("Aaas","Aacs"))
prot_boxplot
```

# Differential analysis

```{r differential_analysis}
formule_contrast <- list("A2016vsWT"="A2016T-WT")

differential_results <- differential_analysis(proteome_data = normalized_data, 
                                              formule_contrast = formule_contrast, 
                                              fc_thr=0.75, pval_fdr = "p_val", pval_thr=0.05, signal_thr=0)

differential_results

save_differential_analysis_table(proteome_data = normalized_data, 
                                 differential_results = differential_results, 
                                 dirOutput=outputdir)

ploft_diff_number <- generate_differential_barplots(differential_results, 
                                                    data_type="protein")
ploft_diff_number

ploft_diff_number <- generate_differential_barplots(differential_results, 
                                                    data_type="peptide")
ploft_diff_number
```

# Test vulcano plot

```{r vulcano_plot}
generate_volcano_plots_protein<-generate_volcano_plots(differential_results, 
                                                       data_type="protein", comparison="A2016vsWT")
generate_volcano_plots_protein

generate_volcano_plots_peptide<-generate_volcano_plots(differential_results, 
                                                       data_type="peptide", comparison="A2016vsWT")
generate_volcano_plots_peptide
```


# MDS differential analysis

```{r mds_differential}
mds_diff <- mds_differential_analysis_plot(differential_analysis = differential_results, 
                                           proteome_data = normalized_data, 
                                           type = "protein")
mds_diff

mds_diff_pep <- mds_differential_analysis_plot(differential_analysis = differential_results, 
                                               proteome_data = normalized_data, 
                                               type = "peptide")
mds_diff_pep
```

# pca differential analysis

```{r pca_differential}
pca_diff <- pca_differential_analysis_plot(differential_analysis = differential_results, 
                                           proteome_data = normalized_data, 
                                           type = "protein")
pca_diff

pca_diff_pep <- pca_differential_analysis_plot(differential_analysis = differential_results, 
                                               proteome_data = normalized_data, 
                                               type = "peptide")
pca_diff_pep
```

# TEST enrichement

```{r enrichment}
enrichemnt_results <- perform_enrichment_analysis(differential_results = differential_results, 
                                                  enrichR_custom_DB = T,
                                                  enrich_filter_DBs=c("Human_Gene_Atlas","BioPlanet_2019",
                                                                      "KEGG_2021_Human",
                                                                      "Reactome_2022","ChEA_2022"), 
                                                  dirOutput=outputdir)

plots_all <- enrichment_figure(enr_df = enrichemnt_results,
                               category = c("all"), 
                               save=T, 
                               dirOutput=outputdir)
plots_all

plots_down <- enrichment_figure(enr_df = enrichemnt_results, 
                                category = c("down","up"),
                                save=F)
plots_down

plots_terms <- enrichment_figure(enr_df = enrichemnt_results, 
                                 category = c("down","up"), 
                                 save=F, 
                                 enrich_filter_term = c("cell","pathways"))
plots_terms
```

# TEST UNIVERSE ENRICHMENT

```{r universe_enrichment}
# TODO: add universe enrichment
```

# TEST network string

```{r string network}

strindb_res <- STRINGdb_network(differential_results = differential_results, 
                                species="Homo sapiens", 
                                dirOutput=outputdir)
strindb_res  

```

# TEST KINASE TREE

```{r kinase_tree}
kinase_res <- kinase_tree(proteome_data = normalized_data, 
                          differential_results = differential_results, 
                          formule_CORAL = formule_contrast, 
                          dirOutput=outputdir)
kinase_res
```

