---
title: "interactn_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{interactn_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Load InteracTN and extract example
```{r setup}
library(proTN)
extract_example(path_interactn = tempdir())
```

## Read example data

Raw abundances are determined from the area under the peak (m/z, time, number of events) for each PSM (Peptide-to-Spectrum Match).

Initial filtering removes the following peptides:

-   Low/missing abundance: PSMs not consistently present in all replicates of at least one condition are removed.
-   Contaminants.

After filtering, raw abundances are log2 transformed.

```{r read_proteomics_MQ}
proteome_data <- read_proteomics(software = "MQ",
                                 folder = paste0(tempdir(),"/case_study_interactn/"),
                                 peptide_filename = "evidence")
```

#### Distribution of log2 raw abundances
Violin plot showing sample distributions of log2 raw abundances of peptides.

```{r preprocess_figure, fig.height=5, fig.width=7}
# Available abundance. The red percentage will be imputed.
generate_abundance <- generate_abundance_plot(proteome_data = proteome_data)
generate_abundance$plot
```

#### Percentage of quantified peptides per sample
Bar plot showing the percentage of peptides with available abundances for each sample.

```{r preprocess_figure_2, fig.height=5, fig.width=7}
# Number of peptides per protein
generate_peptide_distribution <- generate_peptide_distribution_plot(proteome_data = proteome_data)
generate_peptide_distribution$plot
```

#### Imputation function

**Imputation** is performed on peptide and protein abundances with the Bioconductor package [PhosR](https://www.sciencedirect.com/science/article/pii/S221112472100084X). Round imputation is performed in the absence of replicates. ProTN uses two functions from PhosR for imputation: the first imputes missing values for a peptide across replicates within a single condition, the second is a tail-based imputation approach as implemented in Perseus.

```{r impute_function}
imputed_data <- impute_intensity(proteome_data = proteome_data)
```

#### Normalization function

**Normalization:**

**Peptide** abundances are normalized with the equalMedianNormalization function from the Bioconductor package [DEqMS](https://www.bioconductor.org/packages/release/bioc/html/DEqMS.html) (equal medians in all samples). 

**Protein** normalized abundances are calculated with the medianSweeping function from the Bioconductor package [DEqMS](https://www.bioconductor.org/packages/release/bioc/html/DEqMS.html).

```{r normalization_function}
normalized_data <- normalization_ProTN(proteome_data = imputed_data)
```

#### Export normalized data to an Excel file

```{r exporter, eval=FALSE}
save_abundance_tables(proteome_data = normalized_data, 
                      dirOutput=outputdir)
```

#### Distribution of protein normalized abundance
Violin plot showing sample distributions of log2 normalized and imputed abundances.

```{r normalized_plot, fig.height=5, fig.width=7}
# Abundance per Protein
prot_plot <- plot_abundance_distribution(proteome_data = normalized_data,
                                         type = "protein")
prot_plot$plot
```

#### Distribution of peptide normalized abundance
Violin plot showing sample distributions of log2 normalized and imputed abundances.

```{r normalized_plot_2, fig.height=5, fig.width=7}
# Abundance per Peptide
pep_plot <- plot_abundance_distribution(proteome_data = normalized_data,
                                        type = "peptide")
pep_plot$plot
```

#### MDS based on protein normalized data
Multidimensional scaling (MDS) takes a set of dissimilarities and returns a set of points such that the distances between the points are approximately equal to the dissimilarities. Classical multidimensional scaling is also known as principal coordinates analysis.

```{r mds_normalized, fig.height=5, fig.width=7}
# Abundance per Protein
prot_mds <- mds_plot(proteome_data = normalized_data,
                     type = "protein")
prot_mds$plot
```

#### MDS based on peptide normalized data
Multidimensional scaling (MDS) takes a set of dissimilarities and returns a set of points such that the distances between the points are approximately equal to the dissimilarities. Classical multidimensional scaling is also known as principal coordinates analysis.

```{r mds_normalized_2, fig.height=5, fig.width=7}
# Abundance per Peptide
pep_mds <- mds_plot(proteome_data = normalized_data,
                    type = "peptide")
pep_mds$plot
```

#### PCA based on protein normalized data
Principal component analysis (PCA) is performed on normalized protein and peptide abundances.

```{r pca_normalized, fig.height=5, fig.width=7}
# Abundance per Protein
prot_pca <- pca_plot(proteome_data = normalized_data,
                     type = "protein")
prot_pca$plot
```

#### PCA based on peptide normalized data
Principal component analysis (PCA) is performed on normalized protein and peptide abundances. 

```{r pca_normalized_2, fig.height=5, fig.width=7}
# Abundance per Peptide
pep_pca <- pca_plot(proteome_data = normalized_data,
                    type = "peptide")
pep_pca$plot
```

#### Boxplot of abundances for selected proteins
Normalized log2 abundances are displayed for proteins selected in the input. Abundances are grouped by condition.
```{r boxplot_protein, fig.height=5, fig.width=7}
prot_boxplot <- plot_selected_proteins(proteome_data = normalized_data,
                                       list_protein = c("Aaas","Aacs","Zyx"))
prot_boxplot$plot
```

#### Heatmap of abundances for selected proteins
Normalized log2 abundances are displayed for proteins selected in the input.
```{r heatmap_protein, fig.height=7, fig.width=7}
prot_heatmap <- heatmap_selected_proteins(proteome_data = normalized_data,
                                       list_protein = c("Aaas","Aacs","Arsa","Bag2","Cd80",
                                                        "Cdk12","Cdk2","Zyx"))
prot_heatmap$plot
```

## Differential analysis

```{r differential_analysis}
formule_contrast <- list("AM_vs_TL"="AM-TLpool", "PBS_vs_TL"="PBS-TLpool", "AM_vs_PBS"="AM-PBS")

differential_results <- differential_analysis(proteome_data = normalized_data,
                                              formule_contrast = formule_contrast,
                                              fc_thr=0.5,
                                              pval_fdr = "p_adj",
                                              pval_thr=0.05, 
                                              interactomics = TRUE)
```

#### Number of differentially expressed proteins for each comparison

```{r differentiated_protein, fig.height=3, fig.width=7}
ploft_diff_number <- generate_differential_barplots(differential_results,
                                                    data_type="protein", 
                                                    color_contrast = c("AM_vs_TL_+" = "#664069",
                                                                       "PBS_vs_TL_+"= "#d85448",
                                                                       "AM_vs_PBS_+"= "#F08816"))
ploft_diff_number$plot
```

#### Number of differentially expressed peptides for each comparison

```{r differentiated_peptide, fig.height=3, fig.width=7}
ploft_diff_number <- generate_differential_barplots(differential_results,
                                                    data_type="peptide")
ploft_diff_number$plot
```

#### Export differential analysis results to an Excel file

```{r exporter_differential, eval=FALSE}
save_differential_analysis_table(proteome_data = normalized_data,
                                 differential_results = differential_results,
                                 dirOutput=outputdir)
```

#### Volcano plot
The volcano plots are interactive ggplotly objects. Not visible inside vignettes.

```{r vulcano_plot, eval=FALSE}
generate_volcano_plots_protein<-generate_volcano_plots(differential_results,
                                                       data_type="protein",
                                                       comparison="PBS_vs_TL", 
                                                       interactomics = TRUE, 
                                                       pval_fdr = "p_adj", 
                                                       fc_thr = 0.5,
                                                       color_contrast = c("AM_vs_TL_+" = "#664069",
                                                                          "PBS_vs_TL_+"= "#d85448",
                                                                          "AM_vs_PBS_+"= "#F08816"))
generate_volcano_plots_protein

generate_volcano_plots_peptide<-generate_volcano_plots(differential_results,
                                                       data_type="protein",
                                                       comparison="AM_vs_TL", 
                                                       interactomics = TRUE, 
                                                       pval_fdr = "p_adj", 
                                                       fc_thr = 0.5)
generate_volcano_plots_peptide
```

#### MDS based on differentially expressed proteins
Multidimensional scaling (MDS) is performed on normalized abundances of differentially expressed proteins.

```{r mds_differential, fig.height=5, fig.width=7}
mds_diff <- mds_differential_analysis_plot(differential_analysis = differential_results,
                                           proteome_data = normalized_data,
                                           type = "protein")
mds_diff$plot
```

#### MDS based on differentially expressed peptides
Multidimensional scaling (MDS) is performed on normalized abundances of differentially expressed peptides.

```{r mds_differential_2, fig.height=5, fig.width=7}
mds_diff_pep <- mds_differential_analysis_plot(differential_analysis = differential_results,
                                               proteome_data = normalized_data,
                                               type = "peptide")
mds_diff_pep$plot
```

#### PCA based on differentially expressed proteins
Principal component analysis (PCA) is performed on normalized abundances of differentially expressed proteins.

```{r pca_differential, fig.height=5, fig.width=7}
pca_diff <- pca_differential_analysis_plot(differential_analysis = differential_results,
                                           proteome_data = normalized_data,
                                           type = "protein")
pca_diff$plot

```

#### PCA based on differentially expressed peptides
Principal component analysis (PCA) is performed on normalized abundances of differentially expressed peptides.

```{r pca_differential_2, fig.height=5, fig.width=7}
pca_diff_pep <- pca_differential_analysis_plot(differential_analysis = differential_results,
                                               proteome_data = normalized_data,
                                               type = "peptide")
pca_diff_pep$plot
```

## Enrichment analysis with EnrichR
Annotation enrichment analysis is performed using the web-based application [EnrichR](https://maayanlab.cloud/Enrichr/), against an updated collection of 98 databases covering Transcription, Pathways, Ontologies, Diseases, Drugs, Cell Types, and Experimental Datasets. 
For each comparison, enrichment is calculated for up-regulated proteins (UP), down-regulated proteins (DOWN), and their union (ALL).

```{r enrichment, eval=FALSE}
enrichemnt_results <- perform_enrichment_analysis(differential_results = differential_results,
                                                  enrichR_custom_DB = T,
                                                  enrich_filter_DBs=c("Human_Gene_Atlas","BioPlanet_2019",
                                                                      "KEGG_2021_Human",
                                                                      "Reactome_2022","ChEA_2022"),
                                                  dirOutput=outputdir)

## Generate figures for the enrichment analysis.
## ALL: analysis computed on the UP+DOWN regulated proteins
## UP: analysis computed on the UP regulated proteins
## DOWN: analysis computed on the DOWN regulated proteins
## enrich_filter_term: show only terms matching the pattern provided.

plots_all <- enrichment_figure(enr_df = enrichemnt_results,
                               category = c("all"),
                               save=F)

plots_down <- enrichment_figure(enr_df = enrichemnt_results,
                                category = c("down","up"),
                                save=F)

plots_terms <- enrichment_figure(enr_df = enrichemnt_results,
                                 category = c("down","up"),
                                 save=F,
                                 enrich_filter_term = c("cell","pathways"))
```

## Protein-protein interaction network analysis of differentially expressed proteins 
Protein-protein interactions (PPIs) are crucial for understanding cell physiology. PPI analysis of differentially expressed proteins is based on the [STRING](https://string-db.org/) database. Communities within the network are identified with the Clauset-Newman-Moore (CNM) algorithm.

```{r string network, eval=FALSE}

strindb_res <- STRINGdb_network(differential_results = differential_results,
                                species="Homo sapiens", 
                                dirOutput=outputdir)
strindb_res  

```
