---
title: "phosprotn_protn_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{phosprotn_protn_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Load ProTN and extraction example
```{r setup}
library(proTN)
extract_example(path_phospho = tempdir())
```

## Read example data

Raw abundances are determined from the area under the peak (m/z, time, number of events) for each PSM (Peptide-to-Spectrum Match).

Initial filtering removes the following peptides:

-   Low/missing abundance: PSMs not consistently present in all replicates of at least one condition are removed.
-   Contaminants.

After filtering, raw abundances are log2 transformed

```{r read_proteomics_MQ}
proteome_data <- read_phospho_proteome_proteomics(software = "MQ", 
                                                  folder_proteome = paste0(tempdir(),"/case_study_phospho/txt_PROTEOME"),
                                                  folder_phospho = paste0(tempdir(),"/case_study_phospho/txt_Phospho"),
                                                  peptide_proteome_filename = "evidence", 
                                                  peptide_phospho_filename = "evidence", 
                                                  phospho_thr = 0.75)
```

#### Distributions of log2 raw abundances
Violin plot showing sample distributions of log2 raw abundances of peptides.
    phospho_* : figure and data.table for phospho-proteomic data
    proteome_* : figure and data.table for proteomic data

```{r preprocess_figure, fig.height=5, fig.width=7}
# Avalaible abundance. The red percentage will be imputed.
generate_abundance <- generate_abundance_plot(proteome_data = proteome_data, phospho_with_proteome=TRUE)
generate_abundance$phospho_plot
```

#### Percentage of quantified peptides per sample
Barplot showing the percentage of peptides with available abundances for each sample.
    phospho_* : figure and data.table for phospho-proteomic data
    proteome_* : figure and data.table for proteomic data

```{r preprocess_figure_2, fig.height=5, fig.width=7}
# Number peptide per each protein
generate_peptide_distribution <- generate_peptide_distribution_plot(proteome_data = proteome_data,
                                                                    phospho_with_proteome=TRUE)
generate_peptide_distribution$phospho_plot
```

#### Percentage of phosphosite residue
The following graph represent the percentage of the phosphosites located on Serine (S), Throenines (T) and Tyrosines (Y).

```{r preprocess_figure_3, fig.height=3, fig.width=7}
create_phosphosite_fig <- create_phosphosite_plot(proteome_data = proteome_data, 
                                                  software = "MQ")
create_phosphosite_fig$plot
```

#### Impute function

**Imputation** is performed on peptide and protein abundances with the Bioconductor package [PhosR](https://www.sciencedirect.com/science/article/pii/S221112472100084X). Round imputation is performed in absence of replicates. ProTN uses two function of PhosR for the imputation, the first impute the missing values for a peptide across replicates within a single condition, the second is a tail-based imputation approach as implemented in Perseus.

```{r impute_function}
imputed_data <- impute_intensity(proteome_data = proteome_data)
```

#### Normalization function

**Normalization:**

**Peptides** abundances are normalized with the equalMedianNormalization function from the Bioconductor package [DEqMS](<https://www.bioconductor.org/packages/release/bioc/html/DEqMS.html>) (equal medians in all samples). 

**Protein** normalized abundances were calculated with the medianSweeping function from the Bioconductor package [DEqMS](<https://www.bioconductor.org/packages/release/bioc/html/DEqMS.html>).

```{r normalization_function}
normalized_data <- normalization_ProTN(proteome_data = imputed_data)
```

#### Exporter function normalized data to an Excel file

```{r exporter, eval=FALSE}
save_abundance_tables(proteome_data = normalized_data, 
                      dirOutput=outputdir)
```

#### Distribution of protein normalized abundance
Violin plot showing sample distributions of log2 normalized and imputed abundances.

```{r normalized_plot, fig.height=5, fig.width=7}
#Abundance per Protein
prot_plot <- plot_abundance_distribution(proteome_data = normalized_data,
                                         type = "protein")
prot_plot$plot
```

#### Distribution of peptide normalized abundance
Violin plot showing sample distributions of log2 normalized and imputed abundances.

```{r normalized_plot_2, fig.height=5, fig.width=7}
#Abundance per Peptide
pep_plot <- plot_abundance_distribution(proteome_data = normalized_data,
                                        type = "peptide")
pep_plot$plot
```

#### MDS based on protein normalized data
Multidimensional scaling (MDS) takes a set of dissimilarities and returns a set of points such that the distances between the points are approximately equal to the dissimilarities. Classical multidimensional scaling is also known as principal coordinates analysis.

```{r mds_normalized, fig.height=5, fig.width=7}
#Abundance per Protein
prot_mds <- mds_plot(proteome_data = normalized_data,
                     type = "protein")
prot_mds$plot
```

#### MDS based on peptide normalized data
Multidimensional scaling (MDS) takes a set of dissimilarities and returns a set of points such that the distances between the points are approximately equal to the dissimilarities. Classical multidimensional scaling is also known as principal coordinates analysis.

```{r mds_normalized_2, fig.height=5, fig.width=7}
#Abundance per Peptide
pep_mds <- mds_plot(proteome_data = normalized_data,
                    type = "peptide")
pep_mds$plot
```

#### PCA based on protein normalized data
Principal component analysis (PCA) is performed on normalized Protein and Peptide abundances.

```{r pca_normalized, fig.height=5, fig.width=7}
#Abundance per Protein
prot_pca <- pca_plot(proteome_data = normalized_data,
                     type = "protein")
prot_pca$plot
```

#### PCA based on peptide normalized data
Principal component analysis (PCA) is performed on normalized Protein and Peptide abundances. 

```{r pca_normalized_2, fig.height=5, fig.width=7}
#Abundance per Peptide
pep_pca <- pca_plot(proteome_data = normalized_data,
                    type = "peptide")
pep_pca$plot
```

#### Boxplot abundances of selected proteins
Normalized log2 abundances are displayed for proteins selected in the input. Abundances are grouped per condition.
```{r boxplot_protein, fig.height=5, fig.width=7}
prot_boxplot <- plot_selected_proteins(proteome_data = normalized_data,
                                       list_protein = c("Aaas","Aacs","Zyx"))
prot_boxplot$plot
```

#### Heatmap abundances of selected proteins
Normalized log2 abundances are displayed for proteins selected in the input.
```{r heatmap_protein, fig.height=7, fig.width=7}
prot_heatmap <- heatmap_selected_proteins(proteome_data = normalized_data,
                                       list_protein = c("Aaas","Aacs","Arsa","Bag2","Cd80",
                                                        "Cdk12","Cdk2","Zyx"))
prot_heatmap$plot
```

## Differential analysis
Differentially phosphorylated peptides are identified using the Limma Bioconductor package. In each comparison, phospho-peptide changes are scaled for differences in the corresponding protein levels. PhosProTN also perform phosphorylation enrichment analysis for each condition.

```{r differential_analysis}
formule_contrast <- list("A2016vsWT"="A2016T-wt")

differential_results <- differential_analysis(proteome_data = normalized_data,
                                              formule_contrast = formule_contrast,
                                              fc_thr=0.5,
                                              pval_fdr = "p_val",
                                              pval_thr=0.05,
                                              signal_thr=0)
```

#### Number of differentially expressed proteins for each comparison

```{r differentiated_protein, fig.height=3, fig.width=7}
ploft_diff_number_ctrl <- generate_differential_barplots(differential_results, 
                                                         data_type="protein", 
                                                         phospho_ctrl = T, 
                                                         color_contrast = c("grey40","grey60",
                                                                            "grey40","grey60",
                                                                            "#664069","#8A628D"))
ploft_diff_number_ctrl$plot
```

#### Number of differentially expressed peptide for each comparison

```{r differentiated_peptide, fig.height=3, fig.width=7}
ploft_diff_number <- generate_differential_barplots(differential_results,
                                                    data_type="peptide")
ploft_diff_number$plot
```

#### Exporter function differential analysis to an Excel file

```{r exporter_differential, eval=FALSE}
save_differential_analysis_table(proteome_data = normalized_data,
                                 differential_results = differential_results,
                                 dirOutput=outputdir)
```

#### Vulcano plot
The vulcano plot are interactive ggplotly. Not visible inside vignettes.

```{r vulcano_plot, eval=FALSE}
generate_volcano_plots_protein<-generate_volcano_plots(differential_results,
                                                       data_type="protein",
                                                       comparison="A2016vsWT")
generate_volcano_plots_protein

generate_volcano_plots_peptide<-generate_volcano_plots(differential_results,
                                                       data_type="peptide",
                                                       comparison="A2016vsWT")
generate_volcano_plots_peptide
```

#### MDS based on differentially expressed peptide
Multidimensional scaling (MDS) are performed on normalized abundances of differentially expressed peptide.

```{r mds_differential, fig.height=5, fig.width=7}
mds_diff_pep <- mds_differential_analysis_plot(differential_analysis = differential_results,
                                               proteome_data = normalized_data,
                                               type = "peptide")
mds_diff_pep$plot
```

#### PCA based on differentially expressed peptide
Principal component analysis (PCA) are performed on normalized abundances of differentially expressed peptide.

```{r pca_differential, fig.height=5, fig.width=7}
pca_diff_pep <- pca_differential_analysis_plot(differential_analysis = differential_results,
                                               proteome_data = normalized_data,
                                               type = "peptide")
pca_diff_pep$plot
```

## Enrichment analisys with EnrichR
Annotation enrichment analysis is performed using the web-based application [EnrichR](https://maayanlab.cloud/Enrichr/), against an updated collection of 98 databases covering Transcription, Pathways, Ontologies, Diseases, Drugs, Cell Types, Experimental Datasets. 
For each comparison enrichment is calculate over up-regulated proteins (UP), down-regulated proteins (DOWN), and their union (ALL).

```{r enrichment, eval=FALSE}
enrichemnt_results <- perform_enrichment_analysis(differential_results = differential_results,
                                                  enrichR_custom_DB = T,
                                                  enrich_filter_DBs=c("Human_Gene_Atlas","BioPlanet_2019",
                                                                      "KEGG_2021_Human",
                                                                      "Reactome_2022","ChEA_2022"),
                                                  dirOutput=outputdir)

## Generate figure regarding the enrichment analysis.
## ALL: analysis computed on the UP+DOWN regulated proteins
## UP: analysis computed on the UP regulated proteins
## DOWN: analysis computed on the DOWN regulated proteins
## enrich_filter_term: show only terms matching the pattern provided.

plots_all <- enrichment_figure(enr_df = enrichemnt_results,
                               category = c("all"),
                               save=F)

plots_down <- enrichment_figure(enr_df = enrichemnt_results, category = c("down","up"), 
                                save=F, color_contrast = c("grey40","grey60"))

plots_terms <- enrichment_figure(enr_df = enrichemnt_results,
                                 category = c("down","up"),
                                 save=F,
                                 enrich_filter_term = c("cell","pathways"))
```

## Protein-protein interaction network analysis of differentially expressed proteins 
Protein-protein interactions (PPIs) are crucial for understanding cell physiopathology. PPI analysis of differentially expressed proteins is based on the [STRING](https://string-db.org/) database.  Communities within the network are identified with the Clauset-Newman-Moore (CNM) algorithm.

```{r string network, eval=FALSE}

strindb_res <- STRINGdb_network(differential_results = differential_results,
                                species="Homo sapiens", 
                                dirOutput=outputdir)
strindb_res  

```

## Kinase Trees
In this step, PhosProTN identifies alterations in the activity of kinases, based on the differentially phosphorylated sites observed in each contrast. The identification of the altered kinases is perfomed by [PhosR](https://github.com/PYangLab/PhosR) package (Kim et al. 2021), which provide a score of kinase activities based on the comparison formula provided. The activities of the kinases are saved in txt files inside the KinaseTree folder.
In addition, graphical kinase trees are generated using an adapted version of the webtool [CORAL](http://phanstiel-lab.med.unc.edu/CORAL/) (Metz K.S. et al. 2018).
"DISCLAIMER CORAL: The kinome tree plots generated by CORAL make use the data generated by: "Manning et al., Science, 2002", and are based on a figure generated by "Cell Signaling Technology": "https://www.cellsignal.com/contents/science/protein-kinases-introduction/kinases"

```{r kinase_tree, eval=FALSE}
kinase_res <- kinase_tree(proteome_data = normalized_data, 
                          differential_results = differential_results, 
                          formule_CORAL = formule_contrast, 
                          dirOutput = outputdir)
kinase_res
```
