---
title: "phosprotn_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{phosprotn_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Load ProTN and extraction example
```{r setup}
library(proTN)
extract_example(path_phospho = tempdir())
```

# Read example data
```{r read_proteomics_MQ}
proteome_data <- read_phosphoproteomics(software = "MQ",
                                 folder = paste0(tempdir(),"/case_study_phospho/txt_Phospho"),
                                 peptide_filename = "evidence")
```

# Pre-process figure
```{r preprocess_figure, fig.height=5, fig.width=7}
# Avalaible abundance. The red percentage will be imputed.
generate_abundance <- generate_abundance_plot(proteome_data = proteome_data)
generate_abundance$plot
# Number peptide per each protein
generate_peptide_distribution <- generate_peptide_distribution_plot(proteome_data = proteome_data)
generate_peptide_distribution$plot
#Percentage of the phosphorilated site
create_phosphosite_fig <- create_phosphosite_plot(proteome_data = proteome_data,
                                                  software = "MQ")
create_phosphosite_fig$plot
```


# Impute function

```{r impute_function}
imputed_data <- impute_intensity(proteome_data = proteome_data)
```

# Normalization

```{r normalization_function}
normalized_data <- normalization_ProTN(proteome_data = imputed_data)
```

# Save the normalized data in an Excel file

```{r exporter, eval=FALSE}
save_abundance_tables(proteome_data = normalized_data, 
                      dirOutput=outputdir)
```

# Distribution of normalized abundance

```{r normalized_plot, fig.height=5, fig.width=7}
#Abundance per Protein
prot_plot <- plot_abundance_distribution(proteome_data = normalized_data,
                                         type = "protein")
prot_plot$plot
#Abundance per Peptide
pep_plot <- plot_abundance_distribution(proteome_data = normalized_data,
                                        type = "peptide")
pep_plot$plot
```

# MDS figure normalized data

```{r mds_normalized, fig.height=5, fig.width=7}
#Abundance per Protein
prot_mds <- mds_plot(proteome_data = normalized_data,
                     type = "protein")
prot_mds$plot
#Abundance per Peptide
pep_mds <- mds_plot(proteome_data = normalized_data,
                    type = "peptide")
pep_mds$plot
```

# PCA figure normalized data

```{r pca_normalized, fig.height=5, fig.width=7}
#Abundance per Protein
prot_pca <- pca_plot(proteome_data = normalized_data,
                     type = "protein")
prot_pca$plot
#Abundance per Peptide
pep_pca <- pca_plot(proteome_data = normalized_data,
                    type = "peptide")
pep_pca$plot
```

# List protein boxplot

## Boxplot
```{r boxplot_protein, fig.height=5, fig.width=7}
prot_boxplot <- plot_selected_proteins(proteome_data = normalized_data,
                                       list_protein = c("Aaas","Aacs","Zyx"))
prot_boxplot$plot
```

## Heatmap
```{r heatmap_protein, fig.height=7, fig.width=7}
prot_heatmap <- heatmap_selected_proteins(proteome_data = normalized_data,
                                       list_protein = c("Aaas","Aacs","Arsa","Bag2","Cd80",
                                                        "Cdk12","Cdk2","Zyx"))
prot_heatmap$plot
```

# Differential analysis

```{r differential_analysis}
formule_contrast <- list("A2016vsWT"="A2016T-wt")

differential_results <- differential_analysis(proteome_data = normalized_data,
                                              formule_contrast = formule_contrast,
                                              fc_thr=0.5,
                                              pval_fdr = "p_val",
                                              pval_thr=0.05,
                                              signal_thr=0)
```

## Number of differential proteins

```{r differentiated_protein, fig.height=3, fig.width=7}
ploft_diff_number <- generate_differential_barplots(differential_results,
                                                    data_type="protein")
ploft_diff_number$plot
```

## Number of differential peptides

```{r differentiated_peptide, fig.height=3, fig.width=7}
ploft_diff_number <- generate_differential_barplots(differential_results,
                                                    data_type="peptide")
ploft_diff_number$plot
```

```{r exporter_differential, eval=FALSE}
save_differential_analysis_table(proteome_data = normalized_data,
                                 differential_results = differential_results,
                                 dirOutput=outputdir)
```

# Vulcano plot
The vulcano plot are interactive ggplotly. Not visible inside vignettes.

```{r vulcano_plot, eval=FALSE}
generate_volcano_plots_protein<-generate_volcano_plots(differential_results,
                                                       data_type="protein",
                                                       comparison="A2016vsWT")
generate_volcano_plots_protein

generate_volcano_plots_peptide<-generate_volcano_plots(differential_results,
                                                       data_type="peptide",
                                                       comparison="A2016vsWT")
generate_volcano_plots_peptide
```


# MDS of differential analysis

```{r mds_differential, fig.height=5, fig.width=7}
mds_diff <- mds_differential_analysis_plot(differential_analysis = differential_results,
                                           proteome_data = normalized_data,
                                           type = "protein")
mds_diff$plot

mds_diff_pep <- mds_differential_analysis_plot(differential_analysis = differential_results,
                                               proteome_data = normalized_data,
                                               type = "peptide")
mds_diff_pep$plot
```

# PCA of differential analysis

```{r pca_differential, fig.height=5, fig.width=7}
pca_diff <- pca_differential_analysis_plot(differential_analysis = differential_results,
                                           proteome_data = normalized_data,
                                           type = "protein")
pca_diff$plot

pca_diff_pep <- pca_differential_analysis_plot(differential_analysis = differential_results,
                                               proteome_data = normalized_data,
                                               type = "peptide")
pca_diff_pep$plot
```

# Enrichment analisys with EnrichR

```{r enrichment, eval=FALSE}
enrichemnt_results <- perform_enrichment_analysis(differential_results = differential_results,
                                                  enrichR_custom_DB = T,
                                                  enrich_filter_DBs=c("Human_Gene_Atlas","BioPlanet_2019","KEGG_2021_Human",
                                                                      "Reactome_2022","ChEA_2022"),
                                                  dirOutput=outputdir)

plots_all <- enrichment_figure(enr_df = enrichemnt_results,
                               category = c("all"),
                               save=F)
plots_all

plots_down <- enrichment_figure(enr_df = enrichemnt_results,
                                category = c("down","up"),
                                save=F)
plots_down

plots_terms <- enrichment_figure(enr_df = enrichemnt_results,
                                 category = c("down","up"),
                                 save=F,
                                 enrich_filter_term = c("cell","pathways"))
plots_terms
```

# StringDB network

```{r string network, eval=FALSE}

strindb_res <- STRINGdb_network(differential_results = differential_results,
                                species="Homo sapiens", 
                                dirOutput=outputdir)
strindb_res  

```

# Kinase tree with PhosR and CORAL

```{r kinase_tree, eval=FALSE}
kinase_res <- kinase_tree(proteome_data = normalized_data, 
                          differential_results = differential_results, 
                          formule_CORAL = formule_contrast, 
                          dirOutput = outputdir)
kinase_res
```
